 

### 2022-04-29 doc/manuscript/tables_figures/DOYtiming_allyears.Rmd
Create pdf versions of figures in main manuscript

---
title: ''
output:
  bookdown::word_document2: 
    reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    fig_caption: yes
bibliography: references.bib
csl: nature.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\raggedright

**Title:** Warm springs alter timing but not total growth of temperate deciduous trees

**Authors:**

Cameron Dow^1,2^ (Orcid ID : 0000-0002-8365-598X)

Albert Y. Kim^1,3^ (Orcid ID : 0000-0001-7824-306X)

Loïc D’Orangeville^4,5^ (Orcid ID : 0000-0001-7841-7082)

Erika B. Gonzalez-Akre^1^ (Orcid ID : 0000-0001-8305-6672)

Ryan Helcoski^1^ (Orcid ID: 0000-0003-3579-0121)

Valentine Herrmann^1^ (Orcid ID : 0000-0002-4519-481X)

Grant L. Harley^6^ (Orcid ID : 0000-0003-1557-8465)

Justin T. Maxwell^7^ (Orcid ID: 0000-0001-9195-3146)

Ian R. McGregor^1,8^ (Orcid ID: 0000-0002-5763-021X)

William J. McShea^1^ (Orchid ID: 0000-0002-8102-02000)

Sean M. McMahon^9,11^ (Orcid ID : 0000-0001-8302-6908)

Neil Pederson^4^ (Orcid ID : 0000-0003-3830-263X)

Alan J. Tepley^1,10^ (Orcid ID : 0000-0002-5701-9613)

Kristina J. Anderson-Teixeira^1,11^* ( Orcid ID : 0000-0001-8461-9713)

**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian National Zoo & Conservation Biology Institute; Front Royal, VA 22630, USA

2. Department of Forestry and Natural Resources, Purdue University, West Lafayette, Indiana, USA

3. Statistical & Data Sciences; Smith College; Northampton, MA 01063, USA

4. Harvard Forest, Petersham, MA 01366, USA

5. Faculty of Forestry and Environmental Management, University of New Brunswick, Fredericton, NB, E3B 5A3, Canada.

6. Department of Earth and Spatial Sciences, University of Idaho, ID 83844, USA

7. Department of Geography, Indiana University, Bloomington, IN 47405, USA

8. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA

9. Smithsonian Environmental Research Center, Edgewater, MD, USA

10. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

11. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama

*corresponding author: teixeirak@si.edu; +1 540 635 6546

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}

## Target journal: Nature

## Nature articles: 3-4 modest display items

#Display items: listing figures and their number here is helpful for planning, and also allows automatic numbering within text -->easy changing of order.

#Figure numbers
schematic_summary=1 
DOY_timing = 2
tree_rings = 3

# EXTENDED DATA  (this chunk must be repeated in Dow_Extended_Data.Rmd)
##tables
EDT_species="Extended Data Table 1"  
EDT_phenology_parameters="Extended Data Table 2" 
EDT_tree_ring_summary="Extended Data Table 3"

##figures
EDF_map = "Extended Data Figure 1" 
EDF_growth_curves_all = "Extended Data Figure 2"  
EDF_climwin = "Extended Data Figure 3"
EDF_pheno_all = "Extended Data Figures 4-5"
EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
EDF_pheno_Tmax_month = "Extended Data Figure 5"
EDF_tree_ring_figs= "Extended Data Figure 6-7"
EDF_tree_rings_TMIN = "Extended Data Figure 6"
EDF_tree_rings_TMAX_extended = "Extended Data Figure 7"

# Supplementary info
chron_table= "SI Table 1"

###~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## some stats/ 
species_table <- read.csv(file = "tables_figures/species_table.csv")
n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
n_dendro_trees_SCBI <- 119 #doing this the lazy way. summed from species_table rather than coded. 
n_dendro_trees_HF <- 321 #doing this the lazy way. summed from species_table rather than coded. 
n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
n_chronologies <- nrow(chronology_table)  
n_cores_sites <- length(unique(chronology_table[, "Location"]))
n_cores_species <- length(unique(chronology_table[, "Species"])) - 1 #the -1 accounts for the Quercus spp. record (combination of white oak group)
n_chronologies_SCBI <- 12
n_chronologies_HF <- 4
```

\newpage

<!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at 2687 words and 1 modest + 2 large display items. Articles typically have 30-50 references (currently at 51)-->

<!-- Articles start with a fully referenced summary paragraph, ideally of no more than 200 words, which is separate from the main text and avoids numbers, abbreviations, acronyms or measurements unless essential. It is aimed at readers outside the discipline. This summary paragraph should be structured as follows: 2-3 sentences of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards. Please refer to our annotated example  to see how the summary paragraph should be constructed.-->

<!--SUMMARY PARAGRAPH-->

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;--><!--@jeong_phenology_2011;-->@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem diameter growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual ring width was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting projections from global C cycle models[@richardson_terrestrial_2012;@keenan_net_2014], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**

<!--BODY-->

In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
<!--@xu_changes_2021 provides an alternate estimate, which is about half of Harris. The main difference being that Xu includes savannas and woodlands, which have experienced a lot of loss from fires. Given focus on forests here, Harris seems more appropriate. -->
Of this important C sink, ~ `r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021],
<!-- alternate estimate: `r round(.31/.88,2)*100`% (temperate net / global net, (@xu_changes_2021. Table 2))--> 
with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019]. <!-- Pugh et al. 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.--> 
The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@ahlstrom_robustness_2012;@friedlingstein_global_2020;@fatichi_moving_2014].

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[<!--@jeong_phenology_2011;--><!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@keenan_net_2014;@finzi_carbon_2020;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021].
Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

Here, we evaluate how spring temperatures affect the timing, rates, and annual increments of stem diameter growth of temperate deciduous trees across eastern North America. 
To test whether warmer springs advance the timing and extend the duration of stem diameter growth, we used dendrometer band measurements on `r n_dendro_trees` trees across two mid-latitude forests. 
To test whether spring temperatures consistently increased annual radial increments, we analyzed `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests.

## Dendrometer band analysis {-#Dendrometer}

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }

<!--Seasonal milestones for both canopy foliage phenology and stem growth occurred 3-16 days earlier at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).-->
<!--Consistent with the results of Ref[@dorangeville_peak_2022], ring porous species began growing earlier, reaching the *DOY~25~* benchmark earlier (by `r 155-128` days at SCBI and `r 166-144` at Harvard Forest), and their growth was spread over a longer growing season (average *L~pgs~* 22 and 17 days longer at SCBI and Harvard Forest, respectively; Fig. `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`).--> 
<!--Peak growing season length was similar across sites, with *L~pgs~* being, on average, only six days longer at SCBI for ring porous species and less than one day longer for diffuse porous species (`r  EDT_phenology_parameters`).-->

Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (CTW). 
CTW was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
The CTW was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (CTW *T~max~*) used as our independent variable.

![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003). Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`. Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }

For ring- and diffuse- porous species at both sites, warmer CTW *T~max~* resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.5 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).

Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig. `r DOY_timing`a-b), there was no consistent lengthening of *L~pgs~* (Fig. `r schematic_summary`, `r EDF_pheno_all`).

In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were usually weak and inconsistent (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
$\Delta DBH$ displayed a weak positive relationship (+0.003-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).

## Tree-ring analysis {-#Tree}

To understand how annual radial stem growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), plus an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`).
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April *T~max~*.
Across all chronologies, the standardized ring-width index (RWI) was significantly (95% confidence interval did not include 0) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`). 
There were even fewer significant positive correlations with March and May *T~max~*: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
In contrast, RWI was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring- and diffuse-porous species, respectively.
*T~min~* generally exhibited weaker relationships to annual growth than *T~max~*, with few significant correlations between spring *T~min~* and RWI (`r EDF_tree_rings_TMIN`).

To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between RWI and *T~max~* in the previous year (`r EDF_tree_rings_TMAX_extended`).
This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of RWI to previous March or April *T~max~* for 5/142 ring porous chronologies and to previous April or May *T~max~* for 7/66 diffuse porous chronologies.

To test whether the negative effect of summer temperatures might offset an enhancement of growth by warmer spring temperatures, we tested for the joint effects of April and June-July *T~max~* on RWI.
Results were qualitatively similar to the univariate correlations (Fig. `r tree_rings`), with significant (p $\le$ 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).

![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**. Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; `r EDT_species`). Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }

## Discussion {-#Discussion}

<!--summary, setting expectation of extra C-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem diameter growth but have little effect on annual increments (Figs. `r schematic_summary`-`r tree_rings`). 
The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020<!--;@stuble_plant_2021-->] and NEE[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase ANPP~woody~, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress[@delpierre_wood_2016;@zweifel_why_2021;@etzold_number_2022], nutrient limitation[@elmore_earlier_2016], daylength[@etzold_number_2022], or sink saturation[@delpierre_wood_2016;@zani_increased_2020].
This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].

<!--discussion of extra C-->
Combined with widespread observations that warming springs tend to lengthen the season of CO~2~ uptake[@lu_no_2022;@finzi_carbon_2020] and increase net annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
It remains theoretically possible that warm springs could augment ANPP~woody~, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
Extra C fixed in years with warm springs could potentially be allocated to formation of more C-dense wood, either through enhanced cell wall thickening (a process that lags behind stem expansion[@cuny_woody_2015]) or to a higher ratio of high-density latewood to lower-density earlywood.
However, existing evidence indicates that vessel features are most strongly controlled by summer drought stress in the previous (earlywood) or current (latewood) year, while warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013]<!--tardif is essential here-->.
Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or ANPP~woody~.

The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, ANPP~woody~ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be compensated by increased respiration in subsequent years.

<!--.short vs long-term effects-->
It is possible that as spring warming continues, forests will adjust to directional changes in growing season length with an enhancement of ANPP~woody~.
Across latitudinal gradients, warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2022], ANPP~woody~[@banburymorgan_global_2021], and NEE[@churkina_spatial_2005].
Thus, warming spring temperatures are likely to increase the biophysical potential for annual tree growth.
If extra photosynthate made available through a growing difference between GPP and ANPP~woody~ is allocated to functions that relieve limitations on woody growth -- for example, by enhancing nutrient and water acquisition through enhanced allocation to roots[@zohner_how_2021;@liu_phenological_2022] -- it is possible that warming spring temperatures could ultimately increase ANPP~woody~ through indirect mechanisms.
Understanding how warming spring temperatures are influencing C allocation within ecosystems remains an important outstanding question.

<!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results clearly demonstrate that the dominant effects of temperature on deciduous tree growth occur not in the spring, but during the peak growing season of the current or sometimes previous year (Fig.`r tree_rings`, `r EDF_tree_rings_TMAX_extended`), when increased atmospheric demand associated with high temperatures can limit both leaf-level gas exchange and stem growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021]. 
Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to *T~max~* (Fig. `r tree_rings`, `r EDT_tree_ring_summary`).
This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress or high temperatures during the peak growing season[@dorangeville_drought_2018;@helcoski_growing_2019;@anderson-teixeira_joint_2022;@babst_twentieth_2019].
<!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of *T~max~* in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study, given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), our tree-ring analysis does reveal a higher frequency of negative than positive correlations of annual growth to spring temperatures, particularly for ring porous species in cooler climates (Fig. `r tree_rings`, `r EDT_tree_ring_summary`).
Thus, warm springs can have a net negative effect on growth, particularly when water is limiting[@gao_earlier_2022].

<!--conclusions-->
As spring temperatures become increasingly warmer, growing seasons will start earlier. 
However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig. `r schematic_summary`) is unlikely to provide a sustained augmentation of C sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@richardson_terrestrial_2012;@keenan_net_2014;@buermann_widespread_2018].
Rather, the dominant effect of rising temperatures on temperate deciduous forest woody productivity will be a negative effect of high summer temperatures (Fig. `r tree_rings`)[@ahlstrom_robustness_2012], which constitutes a positive feedback to climate change.

# References {-}

# Methods {-} 
<!--Nature guidelines: Full methods, typically not exceeding 3000 words. May be subdivided by bold headings.-->

## Dendrometer band analysis {-} 
<!--Study sites -->

Dendrometer band measurements were collected at SCBI[@bourg_initial_2013] and Harvard Forest[@finzi_carbon_2020;@dorangeville_peak_2022], both part of the Forest Global Earth Observatory (ForestGEO)[@anderson-teixeira_ctfsforestgeo_2015;@davies_forestgeo_2021]. 
SCBI (38.8935° N, 78.1454° W; elevation 273–338 m.a.s.l.) is located in the Blue Ridge Mountains at the northern end of Shenandoah National Park, 5 km south of Front Royal, Virginia. 
The forest is secondary and mixed age, having established in the mid-19th century after conversion from agricultural fields[@bourg_initial_2013]. 
Dominant canopy species within the 25.6 ha ForestGEO plot include tulip poplar (*Liriodendron tulipifera* L.), oaks (*Quercus spp.*), and hickories (*Carya spp.*)[@helcoski_growing_2019]. 
The climate is humid temperate, with 1950-2019 mean annual precipitation of 1018 mm and temperatures averaging 1°C in January and 24°C in July[@anderson-teixeira_joint_2022].
Within the study period for the dendrometer band analysis (2011-2019), spring temperatures (March and April *T~max~*; source: CRU v.4.04[@harris_version_2020]) and summer Standardized Precipitation Evapotranspiration Index (SPEI)[@vicente-serrano_multiscalar_2010] values (4-month value of August) were similar to the average climate prior to the study period (1970-2010).
Specifically, average spring *T~max~* was 16.9$\pm$ 1.4°C (mean $\pm$ 1SD) prior to the study period (range: 14.1-19.9) and 17.6$\pm$ 1.7°C (range: 15.6-20) during the study period, while summer SPEI was -0.06$\pm$ 1.02 (range: -1.7-2.4) prior to the study period and 0.5$\pm$ 0.8 (range: -0.8-1.6) during.

>>>>

---
title: ""
author:
output: 
  pdf_document:
    fig_caption: yes
    #toc: true # table of content true
    #toc_depth: 2  
header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003). Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in Extended Data Figure 4. Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](DOYtiming_allyears.png){ width=100% }
